---
title: "Oasis dataset: MRI and Alzheimers"
author: "Livio Finos"
output: 
  html_document: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Info data

<https://www.kaggle.com/datasets/jboysen/mri-and-alzheimers>

**Context:**

The Open Access Series of Imaging Studies (OASIS) is a project aimed at making MRI data sets of the brain freely available to the scientific community <https://www.kaggle.com/datasets/jboysen/mri-and-alzheimers>.

OASIS is made available by the Washington University Alzheimer's Disease Research Center, Dr. Randy Buckner at the Howard Hughes Medical Institute (HHMI)

**Content:**

Longitudinal MRI Data in Nondemented and Demented Older Adults: This set consists of a longitudinal collection of 150 subjects aged 60 to 96. Each subject was scanned on two or more visits, separated by at least one year for a total of 373 imaging sessions. For each subject, 3 or 4 individual T1-weighted MRI scans obtained in single scan sessions are included. The subjects are all right-handed and include both men and women. 72 of the subjects were characterized as nondemented throughout the study. 64 of the included subjects were characterized as demented at the time of their initial visits and remained so for subsequent scans, including 51 individuals with mild to moderate Alzheimer's disease. Another 14 subjects were characterized as nondemented at the time of their initial visit and were subsequently characterized as demented at a later visit.

**Acknowledgements:**

When publishing findings that benefit from OASIS data, please include the following grant numbers in the acknowledgements section and in the associated Pubmed Central submission: P50 AG05681, P01 AG03991, R01 AG021910, P20 MH071616, U24 RR0213

# Homework 
Provide your LMM code, model summary output, coefficient interpretations, and a brief discussion of results (e.g., does dementia accelerate age-related brain volume loss?).

# A First Look


```{r}
dati=read.csv(file = "data/oasis_longitudinal.csv")

dati$Group=factor(dati$Group,levels=c("Nondemented","Converted","Demented"))

```

```{r}
library(ggplot2)
p <- ggplot(dati,aes(Group,nWBV,color=Group))
p+geom_point(size = 3) +geom_boxplot(alpha=.1) + theme_bw()

p <- ggplot(dati,aes(Age,nWBV,color=Group))
p+geom_point(size = 3) +geom_smooth(method = lm) + theme_bw()


```

